Hydrous melting and partitioning in and above the mantle transition zone: insights from water-rich MgO-SiO$_2$-H$_2$O experiments

### 2015-08-02 MSH.tex
Update for submission

\begin{abstract}
Hydrous melting at high pressures affects the physical properties, dynamics and chemical differentiation of the Earth. However, probing the compositions of hydrous melts at the conditions of the mantle transition zone has traditionally been challenging. In this study, we conducted high pressure multianvil experiments at 13 GPa between 1200 and 1900 $^{\circ}$C to investigate the liquidus in the system MgO-SiO$_2$-H$_2$O. Water-rich starting compositions were created using platinic acid (H$_2$Pt(OH)$_6$) as a novel water source. As MgO:SiO$_2$ ratios decrease, the liquidus curve develops an increasingly pronounced concave-down topology. At slab temperatures, the forsterite- and enstatite-water liquidii have similar water contents. Preliminary experiments on the SiO$_2$-H$_2$O binary reveal a pronounced melting point depression up to about 20 mol\% H$_2$O. This melting point depression exceeds estimates from published models, and requires either significant non-ideality between species in the melt, or distinct differences in the ease of protonation of bridging oxygens. We discuss the implications of this work on the migration of deep-seated slab fluids, and the partitioning of water between nominally anhydrous phases and melts.
\end{abstract}

@@

\begin{figure}[h!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/assembly}
  \caption{The 18/11 octahedral assembly design used in this study.}
  \label{fig:assembly}
\end{figure}

@@

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/Z1058_1400C_13GPa_fo}
  \caption{BSE image of typical run products. Experimental chambers from Experiment Z1058, conducted at 1400 $^{\circ}$C and 13 GPa. Chambers have compositions along the forsterite-water binary, with water contents increasing from left to right.}
  \label{fig:sem}
\end{figure}

@@

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/periclase}
  \caption{The periclase-water phase diagram at 13 GPa, based on current experimental results. Model fits to the data use the melt formulation of \cite{SS1985}, with r=1 and K=K(T) (see discussion in main text). Properties of the anhydrous solid and melt are taken from published literature \citep{SLB2011, DKS2013}. An additional subregular interaction term is required to match the composition and temperature of brucite dehydration.}
  \label{fig:MH}
\end{figure}

@@

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/forsterite}
      \caption{The forsterite-water phase diagram at 13 GPa, based on current experimental results. The thermodynamics of pure forsterite are taken from the literature \citep{SLB2011}, and the properties of the anhydrous melt are from the mixing model described in the text. At high temperatures, the forsterite liquidus is metastable with respect to periclase + melt.}
  \label{fig:foH}
\end{figure}
\clearpage

@@

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/enstatite}
  \caption{The clinoenstatite-water phase diagram at 13 GPa, based on current experimental results. The thermodynamics of pure high pressure clinoenstatite are taken from the literature \citep{SLB2011}, and the properties of the anhydrous melt are from the mixing model described in the text.}
  \label{fig:eoH}
\end{figure}

@@

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/stishovite}
      \caption{The silica-water phase diagram at 13 GPa, based on current experimental results. The thermodynamics of stishovite and the anhydrous melt are described in the text. At high temperatures, the stishovite liquidus is metastable with respect to coesite and melt.}
  \label{fig:SH}
\end{figure}

@@

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/experimental-ternary}
  \caption{A preliminary ternary liquidus diagram at 13 GPa, based on current experimental results. Contours are in $^{\circ}$C, and mineral abbreviations are liquidus fields as follows: per - periclase, fo - forsterite, en - clinoenstatite, stv - stishovite, coe - coesite, phB - phase B, br - brucite, chond - hydroxychondrodite, ice - ice VII.}
  \label{fig:ternary}
\end{figure}

@@

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/fo_wad_rw}
  \caption{Water concentrations in wadsleyite, ringwoodite and melt, modelled at 15 and 20 GPa. Experimental data is taken from three published studies \citep{OMY2000,DDFK2005,LSOK2011}.}
  \label{fig:fo_wad_rw_melt}
\end{figure}

@@

\begin{figure}[ht!]
  \centering
  \includegraphics[width=0.8\textwidth]{figures/wad_fo_partitioning_410}
  \caption{Modelled partition coefficient between forsterite, wadsleyite and melt at 14 GPa, approximately the pressure of the 410 km discontinuity. Note the separate scales for solid/solid and solid/melt partitioning. The difference in melt compositions between this study and that of \cite{DDFK2005} results in much lower $D^{wad/melt}$ coefficients. Error bars on the wadsleyite datapoints are those reported by \cite{DDFK2005}.}
  \label{fig:partitioning_wad_fo}
\end{figure}

\begin{figure}[ht!]
  \centering
  \includegraphics[width=0.8\textwidth]{figures/rw_wad_partitioning_520}
  \caption{Modelled partition coefficient between wadsleyite, ringwoodite and melt at 18 GPa, approximately the pressure of the 520 km discontinuity. Note the separate scales for solid/solid and solid/melt partitioning. Error bars on the ringwoodite data points indicate the range of values reported by \cite{OMY2000}. Error bars on the wadsleyite data points indicate 1 s.d. uncertainties as reported by \cite{LSOK2011}.}
  \label{fig:partitioning_rw_wad}
\end{figure}

\clearpage
\section{Conclusions}

\begin{itemize}
\item We report the first experimental results bracketing the hydrous liquidus at conditions existing at the top of the mantle transition zone.
\item Our results include the first high pressure data on the SiO$_2$-H$_2$O liquidus.
\item The water-induced depression of melting in silica-rich systems greatly exceeds that of an ideal mixing model at high pressures, and requires significant stabilisation of the melt at relatively low water contents.
\item Small temperature increases induce large increases in melt productivity as a result of flat liquidus curves. This is especially pronounced for melts in equilibrium with enstatite and stishovite.
\item We produce the first thermodynamic models for water incorporation into forsterite, wadsleyite and ringwoodite at high pressure, where H$_2$O activity in the melt is much less than one.
\item Solid/melt partition coefficients are well-constrained by the data, and in good agreement with reported mechanisms of water incorporation in the olivine polymorphs.
\item Revised solid/solid partition coefficients suggests that small fraction melting may accompany upwelling of peridotite through the wadsleyite-ringwoodite transition.
\end{itemize}

>>>>

\begin{abstract}
Hydrous melting at high pressures affects the physical properties, dynamics and chemical differentiation of the Earth. However, probing the compositions of hydrous melts at the conditions of the mantle transition zone has traditionally been challenging. In this study, we conducted high pressure multianvil experiments at 13 GPa between 1200 and 1900 $^{\circ}$C to investigate the liquidus in the system MgO-SiO$_2$-H$_2$O. Water-rich starting compositions were created using platinic acid (H$_2$Pt(OH)$_6$) as a novel water source. As MgO:SiO$_2$ ratios decrease, the T-$X_{H_2O}$ liquidus curve develops an increasingly pronounced concave-down topology. The melting point reduction of enstatite and stishovite at low water contents exceeds that predicted by simple ideal models of hydrogen speciation. We discuss the implications of this work on the behaviour of melts in the deep upper mantle and transition zone, and present new models describing the partitioning of water between the olivine polymorphs and associated hydrous melts. 
\end{abstract}

@@

\begin{figure}[h!]
  \centering
      %\includegraphics[width=0.8\textwidth]{figures/assembly}
  \caption{The 18/11 octahedral assembly design used in this study.}
  \label{fig:assembly}
\end{figure}

@@

\begin{figure}[ht!]
  \centering
      %\includegraphics[width=0.8\textwidth]{figures/Z1058_1400C_13GPa_fo}
  \caption{BSE image of typical run products. Experimental chambers from Experiment Z1058, conducted at 1400 $^{\circ}$C and 13 GPa. Chambers have compositions along the forsterite-water binary, with water contents increasing from left to right.}
  \label{fig:sem}
\end{figure}

@@

\begin{figure}[ht!]
  \centering
      %\includegraphics[width=0.8\textwidth]{figures/periclase}
  \caption{The periclase-water phase diagram at 13 GPa, based on current experimental results. Model fits to the data use the melt formulation of \cite{SS1985}, with r=1 and K=K(T) (see discussion in main text). Properties of the anhydrous solid and melt are taken from published literature \citep{SLB2011, DKS2013}. An additional subregular interaction term is required to match the composition and temperature of brucite dehydration.}
  \label{fig:MH}
\end{figure}

@@

\begin{figure}[ht!]
  \centering
      %\includegraphics[width=0.8\textwidth]{figures/forsterite}
      \caption{The forsterite-water phase diagram at 13 GPa, based on current experimental results. The thermodynamics of pure forsterite are taken from the literature \citep{SLB2011}, and the properties of the anhydrous melt are from the mixing model described in the text. At high temperatures, the forsterite liquidus is metastable with respect to periclase + melt.}
  \label{fig:foH}
\end{figure}
\clearpage

@@

\begin{figure}[ht!]
  \centering
      %\includegraphics[width=0.8\textwidth]{figures/enstatite}
  \caption{The clinoenstatite-water phase diagram at 13 GPa, based on current experimental results. The thermodynamics of pure high pressure clinoenstatite are taken from the literature \citep{SLB2011}, and the properties of the anhydrous melt are from the mixing model described in the text.}
  \label{fig:eoH}
\end{figure}

@@

\begin{figure}[ht!]
  \centering
      %\includegraphics[width=0.8\textwidth]{figures/stishovite}
      \caption{The silica-water phase diagram at 13 GPa, based on current experimental results. The thermodynamics of stishovite and the anhydrous melt are described in the text. At high temperatures, the stishovite liquidus is metastable with respect to coesite and melt.}
  \label{fig:SH}
\end{figure}

@@

\begin{figure}[ht!]
  \centering
      %\includegraphics[width=0.8\textwidth]{figures/experimental-ternary}
  \caption{A preliminary ternary liquidus diagram at 13 GPa, based on current experimental results. Contours are in $^{\circ}$C, and mineral abbreviations are liquidus fields as follows: per - periclase, fo - forsterite, en - clinoenstatite, stv - stishovite, coe - coesite, phB - phase B, br - brucite, chond - hydroxychondrodite, ice - ice VII.}
  \label{fig:ternary}
\end{figure}

@@

\begin{figure}[ht!]
  \centering
      %\includegraphics[width=0.8\textwidth]{figures/fo_wad_rw}
  \caption{Water concentrations in wadsleyite, ringwoodite and melt, modelled at 15 and 20 GPa. Experimental data is taken from three published studies \citep{OMY2000,DDFK2005,LSOK2011}.}
  \label{fig:fo_wad_rw_melt}
\end{figure}

@@

\begin{figure}[ht!]
  \centering
  %\includegraphics[width=0.8\textwidth]{figures/wad_fo_partitioning_410}
  \caption{Modelled partition coefficient between forsterite, wadsleyite and melt at 14 GPa, approximately the pressure of the 410 km discontinuity. Note the separate scales for solid/solid and solid/melt partitioning. The difference in melt compositions between this study and that of \cite{DDFK2005} results in much lower $D^{wad/melt}$ coefficients. Error bars on the wadsleyite datapoints are those reported by \cite{DDFK2005}.}
  \label{fig:partitioning_wad_fo}
\end{figure}

\begin{figure}[ht!]
  \centering
  %\includegraphics[width=0.8\textwidth]{figures/rw_wad_partitioning_520}
  \caption{Modelled partition coefficient between wadsleyite, ringwoodite and melt at 18 GPa, approximately the pressure of the 520 km discontinuity. Note the separate scales for solid/solid and solid/melt partitioning. Error bars on the ringwoodite data points indicate the range of values reported by \cite{OMY2000}. Error bars on the wadsleyite data points indicate 1 s.d. uncertainties as reported by \cite{LSOK2011}.}
  \label{fig:partitioning_rw_wad}
\end{figure}

\section{Acknowledgements}
The authors would like to thank Gerti Gollner for acquiring capsule materials, Stefan \"Ubelhack and Heinz Fischer for assembly and capsule cutting and Hubert Schulze for sample preparation. RM was supported by a Humboldt Postdoctoral Fellowship. This study was funded by the ACCRETE project (European Research Council Advanced Grant, contract number 290568).
